### 2019-03-03 chapter_00_open_science.tex
Cleanup and add todos

This chapter presents FluidDyn, a project and an ecosystem of packages to
facilitate open-science and open-source in the field of fluid mechanics.  The
chapter is organized into several parts to present the context and describes
our motivations and the concept behind the project FluidDyn as follows:

\begin{itemize}
\item A scientific project based on the collaboration through open-source, such
as FluidDyn, is only possible thanks to the recent developments in methods and
tools for open-source software engineering. We start by presenting these
developments and discuss the implications for the scientists.

\item FluidDyn is based on the language Python. Incidentally, it is the defining
characteristics of Python which make the eventual success of such a
project realizable.
%
Python is one of the most important tool in recent open-source dynamics and in
particular in science.
%
Python in fluid mechanics is growing fast, but is yet to reach widespread
adoption as the language of choice. Therefore, it is useful to present its
characteristics for readers who are unacquainted with its nuances.

@@

Computational sciences, and especially computational fluid dynamics (CFD) as a
discipline has flourished a lot in the latter half of the twentieth century.
Incidentally the origin of this domain can be traced back to the ideas
in \citet{richardson_weather_1922} to achieve numerical weather prediction.
With the arrival of first computer general-purpose computer, ENIAC, the first
simulations were made using a simple barotropic model to make 24 hour
forecasts \citep{lynch_richardson_2010}. In those early days when the use and
availability of computers were limited, such methods were neither widespread
nor feasible. Scientific investigations were dominated by theoretical and
experimental methods.
We have come a long way since then and today one can say that computational
methods now complement theoretical and experimental studies and are equally
important.
Thus to recreate a particular scientific study, the mathematical representation
or even an algorithm leaves a lot to be desired. With the connectivity
provided by the world wide web it is possible to do more and reinvent the way
we perform sciences.

@@

In recent years there has been a move towards the open science
\citep{royal_society_great_britain_science_2012}. Depicted in figure
\ref{fig:opensci} are some of steps needed to ensure reproducibility and
transparency in sciences. \emph{Knowledge} has been traditionally disseminated
in the form of courses and books from universities, and to the wider scientific
community through research articles, conference proceedings and workshops.
While this is nothing new, knowledge has become more accessible to the public
through public-domain and Creative Commons (or similar) licensed course
materials and massively open online courses (MOOCs). Moreover, there is also an
emerging push to publish in open-access journals either through systematic
reforms in academia on how researchers are assessed on their productivity or
through regulations and offering incentives, such as waiving the application
processing charge for publishing \citep{nosek_promoting_2015}.

As mentioned earlier there is a lot of work involved in condensing and
implicating this knowledge into working applications. The \emph{know-how} can
be made accessible and open to scrutiny by making research software open-source
with an appropriate license. To make the codes usable it has to be complemented
by documentation which typically includes tutorials, examples and a detailed
commentary of different components of the code. Furthermore, to inspire
confidence in and to ensure reliability of the code throughout the process of
software development, it needs unit-tests with which a code is continuously
deployed, built and tested --- a process known as Continuous Integration (CI),
which is described in detail later in this chapter. Depending on how much
matured a software is, these can be easy or cumbersome to implement, but
nevertheless, the result is a cleaner and reliable code.

Finally, the \emph{results} generated from the such codes together with the
workflow (i.e. scripts to run the codes and to postprocess the data) when made
open in the form of datasets can ensure that research would be reproducible and
preserved in archives for several years to come \citep{gewin_data_2016}. Open
datasets can be shared and made citeable through services such as
\href{https://zenodo.org}{Zenodo} and \href{https://figshare.com}{Figshare}.

@@

%todo: DONE
% Here, we have to add a paragraph on free and open-source (libre)
% software:

% The following facts can be mentioned:

% - tex (started in 1977), latex (1985)

% - GNU (started 1983), FSF in 1985, licence and philosophy...
% first complete system (with the Linux kernel) in 1991

@@

\paragraphbf{Websites to share knowledge} Mutual aid and knowledge sharing
have been important aspects of the web since its inception and these practices
are now of foremost importance for many human activities.
%
In particular, learning programming and computer science is now greatly aided
by websites like the community driven encyclopedia
\href{https://www.wikipedia.org/}{Wikipedia} or the community driven forum
\href{https://stackoverflow.com/}{Stack Overflow}, where people write thousands
of questions and answers on programming each week.
% todo: remark from Julien Salort: it's old.
% av: So are technologies like git and hg, but we are putting together what
% characterizes our workflow, aren't we? We can emphasize that these forums
% often work only when there is a "thriving community" like the Python
% userbase.
IRC protocol has continued to sustain the test of time. Instant messaging
channels such as \codeinline{\#python} on Freenode facilitate interactive
discussions and a form of community-level support system --- a role which used
to be filled by mailing lists alone in the past. Alternatives such as Riot
(Matrix protocol), Gitter, and Slack have gained traction offering different
integrations solutions on top of instant messaging.
%
Through instant messaging a geographically separated team, as is the case with
FluidDyn project, can achieve near real-time collaboration without the hassle
of setting up video conferences or meetings.

@@

% todo: DONE
% We need an introductory paragraph on how big is Python today
% + the fact that several scientists miss this fact.

% See the file ./python-one-main-languages.rst

@@

\paragraphbf{Ahead Of Time compilation}  Since Python is primarily a ``glue''
language, a common solution is to use code written in C, C++ or Fortran.
%
% todo: DONE
% add a reference to ctypes and cffi to call C with pure Python code
%
It is possible, without a compilation step, to call functions in C libraries
through pure python code using \pack{ctypes} standard module or the \pack{cffi}
package.
%
Libraries with Python bindings can also be used directly (for e.g., wrapped
using SWIG or Boost.Python) or one has to write extensions, i.e.\ Python
modules written in C or C++ using the CPython API that can be imported directly
from Python like any Python module.
%
However, to rely on such techniques to call C libraries would inhibit
development of extending functionalities, as they have a steeper learning
curve compared to Python.
%
Nevertheless, this strategy has given rise to the base modules of the scientific
Python stack, namely \Numpy, \Scipy and \pack{Matplotlib}.

@@

% todo: DONE
% pa: I am really uncomfortable with this presentation which gives the
% impression that we think that the arguments against sharing are always wrong.
%
% We have to make it clear that it is not the case and that we do respect how
% other people think and work.
% av: I have commented out specific counter arguments, instead added two
% sentences after the bullets
\begin{itemize}
\item \textit{``We do not share to keep a comparative advantage.''}

@@

\section{FluidDyn project}
%: Breaking the status quo}
% \section{Implementation and architecture}

We have seen that there is a strong dynamics in play around the use of
computers (in particular with the web) and that this creates very efficient
tools and methods for collective work and software development.
%
The potential of these tools is not fully exploited yet, in the field of fluid
mechanics.

@@

% FluidDyn is an attempt to set off collaborative dynamics based on open-source
% development in fluid dynamics research.
% %
% We shall try, with this project, to explore the possibilities of open-source in
% science and fluid dynamics, by fully exploiting the new open-source tools and
% methods presented in the first subsections of the introduction.
%
% % todo: again mention some success of community-driven subject-area research
% % libraries? See the file ./communitydriven_research_soft.md
%
% The project is right now in a very preliminary stage.  Packages are actively
% evolving with interesting features and a framework for collaborative
% development, packaging, documenting and testing is now well set.
% %
% However, the community around the project is currently tiny and now, we have to
% work on attracting users and developers since an active community is a
% criteria for success and sustenance of an open-source project.
%
% Will people use the FluidDyn tools and collaborate through the project
% FluidDyn?
% %
% There are clearly many challenges and potential barriers:
% \begin{itemize}
% \item Some habits in the community.
% \item Knowledge and skills in the community. For example, only few people use
% issue tracker and pull requests.
% % from Julien Salort: bad!
% % \item Misunderstanding between colleagues and in particular colleagues with
% % responsibilities and power.
% \item Lack of a business model for open-source software in science.  A good
% quality open-source software has a cost that institutions should be willing
% to fund and support.
% \item Lack of recognition of the work spent in open-source, in particular for
% precarious scientists.
% \end{itemize}
%
% On our side, we also have positive points.
% %
% The quality of the tools we use (Python and its scientific ecosystem,
% Mercurial, Read the Docs, Jupyter, ...) is impressive.
% %
% Scientific code is done to be read and to transmit ideas. To this effect, Python
% is among the best languages today.
% %
% We can benefit from the dynamics of Python and of emerging subjects like deep
% learning and the Internet of Things.
% %
% Finally, if we manage to gather a community of users and of developers, the
% collective efficiency related to open-source methods and tools can be a strong
% booster.

>>>>

This chapter presents FluidDyn, a project and an ecosystem of packages to
facilitate open-science and open-source development in the field of fluid
mechanics.
%
The chapter is organized into several parts to present the context and to
describe our motivations and the concept behind the project FluidDyn as
follows:

\begin{itemize}
\item A scientific project based on the collaboration through open-source, such
as FluidDyn, is only possible thanks to the recent developments in methods and
tools for open-source software engineering. We start by presenting these
developments and discuss the implications for scientists.

\item FluidDyn is based on the language Python. Incidentally, it is the defining
characteristics of Python which make the eventual success of such a
project realizable.
%
Python is one of the most important tool in recent open-source dynamics,
particularly in science.
%
Python in fluid mechanics is growing fast, but is yet to reach widespread
adoption as the language of choice. Therefore, it is useful to present its
characteristics for readers who are unacquainted with its nuances.

@@

Computational sciences, and especially computational fluid dynamics (CFD) as a
discipline has flourished a lot in the latter half of the twentieth century.
Incidentally the origin of this domain can be traced back to the ideas
in \citet{richardson_weather_1922} to achieve numerical weather prediction.
With the arrival of first electronic general-purpose computer, ENIAC, the first
simulations were made using a simple barotropic model to make 24 hour
forecasts \citep{lynch_richardson_2010}. In those early days when the use and
availability of computers were limited, such methods were neither widespread
nor feasible. Scientific investigations were dominated by theoretical and
experimental methods.
We have come a long way since then and today one can say that, computational
methods now complement theoretical and experimental studies, and are equally
important.
Thus to recreate a particular scientific study, the mathematical representation
or even an algorithm leaves a lot to be desired. With the connectivity
provided by the world wide web it is possible to do more and reinvent the way
we perform sciences.

@@

In recent years there has been a move towards the open science
\citep[see the report by][]{royal_society_great_britain_science_2012}.
%
Depicted in figure \ref{fig:opensci} are some of steps needed to ensure
reproducibility and transparency in sciences.
%
\emph{Knowledge} has been traditionally disseminated in the form of courses and
books from universities, and to the wider scientific community through research
articles, conference proceedings and workshops.

While this is nothing new, knowledge has become more accessible to the public
through public-domain and Creative Commons (or similar) licensed course
materials and massively open online courses (MOOCs). Moreover, there is also an
emerging push to publish in open-access journals\footnote{In the European
  Union, a proposal to implement Plan S (\url{https://www.coalition-s.org}) was
  been set in motion in September 2018 to fast track adoption of open-access.
} either through systematic
reforms in academia on how researchers are assessed on their productivity or
through regulations and offering incentives, such as waiving the application
processing charge for publishing \citep{nosek_promoting_2015}.

As mentioned earlier there is a lot of work involved in condensing and
implementing this knowledge in the form of working applications.
%
The \emph{know-how} can be made accessible and open to scrutiny by making
research software open-source with an appropriate license.  To make the codes
usable it has to be complemented by documentation which typically includes
tutorials, examples and a detailed commentary of different components of the code.
%
Furthermore, to inspire confidence in and to ensure reliability of the code
throughout the process of software development, it needs unit-tests with
which a code is continuously deployed, built and tested --- a process known as
Continuous Integration (CI), which is described in detail later in this
chapter. Depending on how much matured a software is, these can be easy or
cumbersome to implement, but nevertheless, the result is a cleaner and reliable
code.

Finally, the \emph{results} generated from the such codes together with the
workflow (i.e. scripts to run the codes and to postprocess the data) when made
open in the form of datasets can ensure that research would be reproducible and
preserved in archives for several years to come \citep{gewin_data_2016}. Open
datasets can be shared and made citable through services such as
\href{https://zenodo.org}{Zenodo} and \href{https://figshare.com}{Figshare}.

@@

@@

\paragraphbf{Websites to share knowledge} Mutual aid and knowledge sharing
have been important aspects of the web since its inception and these practices
are now of foremost importance for many human activities.
%
In particular, learning programming and computer science is now greatly aided
by websites like the community driven encyclopedia
\href{https://www.wikipedia.org/}{Wikipedia} or the community driven forum
\href{https://stackoverflow.com/}{Stack Overflow}, where people write thousands
of questions and answers on programming each week.
IRC protocol has continued to sustain the test of time. Instant messaging
channels such as \codeinline{\#python} on Freenode facilitate interactive
discussions and a form of community-level support system --- a role which used
to be filled by mailing lists alone in the past. Alternatives such as Riot
(Matrix protocol), Gitter, and Slack have gained traction offering different
integrations solutions on top of instant messaging.
%
Through instant messaging a geographically separated team, as is the case with
FluidDyn project, can achieve near real-time collaboration without the hassle
of setting up video conferences or meetings.

@@

@@

\paragraphbf{Ahead Of Time compilation}  Since Python is primarily a ``glue''
language, a common solution is to use code written in C, C++ or Fortran.
%
It is possible, without a compilation step, to call functions in C libraries
through pure python code using \pack{ctypes} standard module or the \pack{cffi}
package.
%
Libraries with Python bindings can also be used directly (for e.g., wrapped
using SWIG or Boost.Python) or one has to write extensions, i.e.\ Python
modules written in C or C++ using the CPython API that can be imported directly
from Python like any Python module.
%
However, to rely on such techniques to call C libraries would inhibit
development of extending functionalities, as they have a steeper learning
curve compared to Python.
%
Nevertheless, this strategy has given rise to the base modules of the scientific
Python stack, namely \Numpy, \Scipy and \pack{Matplotlib}.

@@

% TODO: perhaps include the commented out sections in the beginning of the chapter
% pa: I am really uncomfortable with this presentation which gives the
% impression that we think that the arguments against sharing are always wrong.
%
% We have to make it clear that it is not the case and that we do respect how
% other people think and work.
% av: I have commented out specific counter arguments, instead added two
% sentences after the bullets
\begin{itemize}
\item \textit{``We do not share to keep a comparative advantage.''}

@@

\section{FluidDyn project}
%: Breaking the status quo}
% \section{Implementation and architecture}
% TODO: a lot of self-plagiarism below - needs rewrite
We have seen that there is a strong dynamics in play around the use of
computers (in particular with the web) and that this creates very efficient
tools and methods for collective work and software development.
%
The potential of these tools is not fully exploited yet, in the field of fluid
mechanics.

@@

% FluidDyn is an attempt to set off collaborative dynamics based on open-source
% development in fluid dynamics research.
% %
% We shall try, with this project, to explore the possibilities of open-source in
% science and fluid dynamics, by fully exploiting the new open-source tools and
% methods presented in the first subsections of the introduction.
%
% % again mention some success of community-driven subject-area research
% % libraries? See the file ./communitydriven_research_soft.md
%
% The project is right now in a very preliminary stage.  Packages are actively
% evolving with interesting features and a framework for collaborative
% development, packaging, documenting and testing is now well set.
% %
% However, the community around the project is currently tiny and now, we have to
% work on attracting users and developers since an active community is a
% criteria for success and sustenance of an open-source project.
%
% Will people use the FluidDyn tools and collaborate through the project
% FluidDyn?
% %
% There are clearly many challenges and potential barriers:
% \begin{itemize}
% \item Some habits in the community.
% \item Knowledge and skills in the community. For example, only few people use
% issue tracker and pull requests.
% % from Julien Salort: bad!
% % \item Misunderstanding between colleagues and in particular colleagues with
% % responsibilities and power.
% \item Lack of a business model for open-source software in science.  A good
% quality open-source software has a cost that institutions should be willing
% to fund and support.
% \item Lack of recognition of the work spent in open-source, in particular for
% precarious scientists.
% \end{itemize}
%
% On our side, we also have positive points.
% %
% The quality of the tools we use (Python and its scientific ecosystem,
% Mercurial, Read the Docs, Jupyter, ...) is impressive.
% %
% Scientific code is done to be read and to transmit ideas. To this effect, Python
% is among the best languages today.
% %
% We can benefit from the dynamics of Python and of emerging subjects like deep
% learning and the Internet of Things.
% %
% Finally, if we manage to gather a community of users and of developers, the
% collective efficiency related to open-source methods and tools can be a strong
% booster.
